Spfy: distributed predictive genomics of E.coli with graph based result linkage

### 2017-08-11 paper-webserver.tex
UPDATE: moved old material into relevant sections, deleted irrelevant material

% end of new intro
% below material is just old stuff, might or might not be used elsewhere

% costs of bacterial infections
% We need to state the "big problem" up-front that we are working on, which is not really tracking foodborne illness. It is fast analyses of bacterial whole-genomes, providing reference-laboratory tests from in-silico data, predicting phenotype from genotype, and providing a linked database of known genomes that don't need to be re-computed. This can be used in the tracking of foodborne illness, but is more an application of the problem we are working on.
The isolation and characterization of bacterial pathogens are critical for Public Health laboratories to rapidly respond to outbreaks, and to effectively monitor known and emerging pathogens through surveillance programs.
Within Canada it is estimated that over 4,000 hospitalizations and 100 deaths occur due to bacterial infections \cite{thomas2015estimates}; in the United States, over 55,000 hospitalizations and 1,300 deaths occur per year \cite{scallan2011foodborne}. Within North America the total economic burden of these illnesses is estimated to be ACTUAL NUMBER \cite{todd1989costs,scharff2010health}.
The bacterial species that are most frequently associated with severe disease, and cause the majority of this health burden in (Canada / NA/ Europe / Worldwide ) are ABC.
%You could combine the Canada / US numbers into a North America number, and include world-wide data as well. Always avoid vacuous descriptors like "many", "various" -- use an actual quantifier. Eg. Most people prefer vanilla to chocolate. Fifty-one percent of people prefer vanilla to chocolate. Ninety-nine percent of people prefer vanilla to chocolate.

% routinely monitor emerging threats, of which \textit{Salmonella enterica} and \textit{Escherichia coli} are amongst the two most common \cite{kozak2013foodborne}. In \textit{Salmonella enterica} \cite{bell2016recent}, \textit{Escherichia coli} \cite{fratamico2016advances}, and (others pathogens --> either state the other pathogens or leave it out entirely) \cite{ronholm2016navigating}

% surveillance
Until recently, Public-health agencies relied on laboratory tests such as XYZ to characterize bacterial isolates in outbreak and surveillance settings.
The previous gold-standard in determining strain relatedness was pulsed-field gel electrophoresis (PFGE) {ronholm2016navigating}, which uses rare-cutting restriction enzymes to produce a unique banding pattern for each strain.
However, in \textit{Enterococcus faecium}, PFGE has been shown to misclassify 9 of 132 isolates, when compared to whole-genome sequencing (WGS) based discrimination \cite{pinholt2015multiple}.
In \textit{Klebsiella pneumoniae} \cite{marsh2015genomic}, \textit{Yersinia enterocolitica} \cite{gilpin2014limitations}, and \textit{Staphylococcus aureus} \cite{doi:10.1093/ofid/ofu096}, WGS was used to discriminate isolates after initial clustering by PFGE resulted in indistinguishable samples.
Examination of PFGE bands are also subjective, difficult to share \cite{lytsy2017time}, and collative platforms such as PulseNet reported \cite{gilpin2014limitations} that even after collecting 72\% of \textit{Campylobacter jejuni} in a given year in Minnesota (673 cases), 87\% of isolates could not be linked by PFGE pattern.
Antimicrobial resistance testing, virulence factor testing ...
However, current efforts are focused on predictive genomics, where the relevant phenotypic information can be determined through examination of the whole-genome sequence.

% potential of WGS
, and as such can be used to evaluate the spread of outbreaks with better resolution and context than traditional methods \cite{ronholm2016navigating}.
However, while these WGS-based methods can fully examine the underlying genetics, the expressed phenotype is more difficult to predict.
For example, predictions involving reference databases are only as good as the databases themselves.

% drawbacks - big data
Many previous bioinformatics software programs have been developed \textit{ad hoc}, without the use of software engineering principles \cite{de2015trends}.
Such tools were often script-based, with custom data formats, and only suitable for small collections of data \cite{de2015trends}.
While this was acceptable for smaller analyses, bioinformatic pipelines utilizing WGS data are larger and involve linked dependencies, which require the application of systems engineering principles \cite{schatz2015biological}.
Additionally, many subsets of biology now require the analyses of big-data, where the ability to perform computations in real-time, store data in flexible databases, and utilize a common application programming interface (API) linking resources are required \cite{swaminathan2016review}.

% Solution via software engineering

In addition, GenomeTrackr from the Food and Drug Administration (FDA) \cite{allard2016practical}; GenBank from the National Center for Biotechnology Information (NCBI) \cite{doi:10.1093/nar/gks1195}; and PulseNet from the Center for Disease Control and Prevention (CDC) \cite{swaminathan2001pulsenet}, all enable sharing of genome sequences and associated metadata, either openly or amongst member institutions.

%By comparing a pre-collected set of reference gene sequences against a WGS result, bioinformatics software predicts wet-lab serotyping results \cite{whiteside2016superphy}, Virulence Factor (VF) results \cite{joensen2014real}, and antimicrobial resistance (AMR) results \cite{mcarthur2013comprehensive}.
%MLST software uses a similar reference-based approach \cite{larsen2012multilocus}.

While these programs are useful for computing novel analyses, storing results in a tiered structure, with rapid access available for retrieval while maintaining long-term data storage \cite{schatz2015biological}, is needed to prevent the constant re-computation of the same data, and to provide pre-computed analyses for common predictive genomic analyses needed by the public health, and research community \cite{de2015trends}.

% why Superphy
Currently, a number of organizations are creating user-friendly web services for common analyses of bacterial genomes.
These include: NAME \cite{joensen2014real,larsen2012multilocus}, NAME \cite{liu2016construction}, and NAME \cite{hasman2015detection}, which do X, Y and Z respectively.

While useful, leveraging big-data results from WGS is difficult due to the lack of bioinformatic standards and common infrastructure technologies.
Integrating results from multiple genomes, and examining shared connections, is an area that currently requires more research \cite{fricke2014bacterial},

% Superphy
, which provided pre-computed, predictive genomic analyses for determining Shiga toxin subtype, AMR gene profiles, the presence of known virulence factors, and provides predictive biomarkers, both the presence / absence of genomic regions and single nucleotide variants, for bacterial sub-groups based on metadata such as serotype, host source, or location.

Lastly, these services are containerized using Docker to mitigate the complexities of the underlying bioinformatic pipelines.
All services are publicly available at \url{https://lfz.corefacility.ca/superphy/spfy/}.

@@

\section{IMPLEMENTATION}
% para
% goals: predictive analylitcs in minutes, storage up 50,000-100,000 genomes {}

@@

\begin{figure}[t]
\begin{center}
\includegraphics{NAR-fig1.eps}
\end{center}
\caption{Caption for figure within column.}
\label{NAR-fig1}
\end{figure}

@@

% e.g.
% database statistics
% analysis run-time / throughput with different levels of parallelization

@@

% Mention where users / developers can find documentation

@@

% para
% focus on application

\subsection{Deployment to Cloud-Computing Platforms}
% how costs can be saved because using standard web tech means you can deploy to different cloud computing services

@@

% namely galaxy

@@

\begin{table}[b]
\tableparts{%
\caption{This is a table caption}
\label{table:01}%
}{%
\begin{tabular*}{\columnwidth}{@{}lllll@{}}
\toprule
Col. head 1 & Col. head 2 & Col. head 3 & Col. head 4 & Col. head 5
\\
& (\%) & (s$^{-1}$) & (\%) & (s$^{-1}$)
\\
\colrule
Row 1 & Row 1 & Row 1 & -- & --
\\
Row 2 & Row 2 & Row 2 & Row 2 & Row 2
\\
\botrule
\end{tabular*}%
}
{This is a table footnote}
\end{table}

\begin{figure*}[t]
\begin{center}
\includegraphics{NAR-fig2.eps}
\end{center}
\caption{Caption for wide figure over two columns.
\textbf{(a)} Left figure.
\textbf{(b)} Right figure (see (a)).
}
\label{NAR-fig2}
\end{figure*}

\section{CONCLUSION}

>>>>

% end of new intro

@@

\section{IMPLEMENTATION}
% para
% goals: predictive analylitcs in minutes, storage up 50,000-100,000 genomes {}
% drawbacks - big data
Many previous bioinformatics software programs have been developed \textit{ad hoc}, without the use of software engineering principles \cite{de2015trends}.
Such tools were often script-based, with custom data formats, and only suitable for small collections of data \cite{de2015trends}.
While this was acceptable for smaller analyses, bioinformatic pipelines utilizing WGS data are larger and involve linked dependencies, which require the application of systems engineering principles \cite{schatz2015biological}.
Additionally, many subsets of biology now require the analyses of big-data, where the ability to perform computations in real-time, store data in flexible databases, and utilize a common application programming interface (API) linking resources are required \cite{swaminathan2016review}.

@@

We developed Spfy using continuous integration (CI) techniques: all changes and update to the codebase is automatically ran against a series of tests ensuring functionality and backwards compatibility.
We chose TravisCI \url{travisci.io} which offers free services for open-source projects.
The individual tests use PyTest.

@@

% e.g.
% database statistics
We tested spfy with 55,353 samples genomes (267GB), storing both the entire genomes and results for all included analyses modules.
The resulting database had XYZ nodes and XYZ edges, with XYZ object properties.
This worked our to XYZ GB of data stored.

% analysis run-time / throughput with different levels of parallelization

@@

% Mention where users / developers can find documentation
A guide to setting up Spfy is provided at \url{https://github.com/superphy/backend} and a developer guide is provided at \url{https://superphy.readthedocs.io/en/latest/}.

@@

% para
% focus on application
The isolation and characterization of bacterial pathogens are critical for Public Health laboratories to rapidly respond to outbreaks, and to effectively monitor known and emerging pathogens through surveillance programs.
Until recently, Public-health agencies relied on laboratory tests such as XYZ to characterize bacterial isolates in outbreak and surveillance settings.
The previous gold-standard in determining strain relatedness was pulsed-field gel electrophoresis (PFGE) {ronholm2016navigating}, which uses rare-cutting restriction enzymes to produce a unique banding pattern for each strain.
However, in \textit{Enterococcus faecium}, PFGE has been shown to misclassify 9 of 132 isolates, when compared to whole-genome sequencing (WGS) based discrimination \cite{pinholt2015multiple}.
In \textit{Klebsiella pneumoniae} \cite{marsh2015genomic}, \textit{Yersinia enterocolitica} \cite{gilpin2014limitations}, and \textit{Staphylococcus aureus} \cite{doi:10.1093/ofid/ofu096}, WGS was used to discriminate isolates after initial clustering by PFGE resulted in indistinguishable samples.
Examination of PFGE bands are also subjective, difficult to share \cite{lytsy2017time}, and collative platforms such as PulseNet reported \cite{gilpin2014limitations} that even after collecting 72\% of \textit{Campylobacter jejuni} in a given year in Minnesota (673 cases), 87\% of isolates could not be linked by PFGE pattern.
Antimicrobial resistance testing, virulence factor testing ...
However, current efforts are focused on predictive genomics, where the relevant phenotypic information can be determined through examination of the whole-genome sequence. 
, and as such can be used to evaluate the spread of outbreaks with better resolution and context than traditional methods \cite{ronholm2016navigating}.

% application: results similar to a wet-lab
Spfy uses WGS results.
After intial sequencing of new isolates, Spfy can be used in place of a traditional reference laborartory, to determine the O-type and H-type, Stx type, and all known VFs and AMR genes in real-time.
These results can be shared with other ageencies and researchers over the internet.
Futhermore, using Spfy's database of pre-processed genomes, Spfy can determine all strains a sample may be related to which is useful for ...
Cost benefits ...

\subsection{Deployment to Cloud-Computing Platforms}
% how costs can be saved because using standard web tech means you can deploy to different cloud computing services
Spfy's common infrastructure allows the platform to be deployed to multiple cloud computing services, without need to modify the code.
Docker containerization is supported by Amazon AWS, Google Cloud, ...
Neglibble impact of Docker containers on performance \cite{di2015impact}.
Costs ...

@@

% namely galaxy
Other pipeline technologies such as Galaxy \cite{goecks2010galaxy}, Kepler ... \cite{de2015trends}.
important to have rapid access available for retrieval while maintaining long-term data storage \cite{schatz2015biological}
long term storage is needed to prevent the constant re-computation of the same data, and to provide pre-computed analyses for common predictive genomic analyses needed by the public health, and research community \cite{de2015trends}.
Integrating results from multiple genomes, and examining shared connections, is an area that currently requires more research \cite{fricke2014bacterial},

@@
